---
title: "Novel Geomorph Terrain Analysis"
author: "Matthew Ross"
date: "6/19/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(raster)
library(sf)
library(mapview)
library(terra)
library(rayshader)
#devtools::install_github("giswqs/whiteboxR")

library(whitebox)
library(tidyverse)


knitr::opts_chunk$set(warning = F, comment = F,
                      message = F, cache=T,
                      fig.width = 8, fig.height = 5)

```


# Datasets and descriptions

This portion of our novel geomorphology project is focused on examining 
shifts in landscapes after human interventions, and answering the question
"Do people have consistent impacts on geomorphic features? If so what are 
these, and why do they happen?" We wanted to keep this analysis open ended 
across landscape types, so we have three landscapes for analysis. 1) Wetlands
impacted by ditching (or not). 2) Piedmont watersheds impacted by historical
agriculture and erosion (or mostly not). 3) Appalachian watersheds pre and post-mining.

## Piedmont watersheds

This data comes from the Calhoun CZO [https://www.hydroshare.org/resource/cbdf5af338214abea9bd3e993aae4d01/] 
and John Mallard. John will write a description here.

### Watershed delineation and processing

To conduct our analyses we need to first prep the DEM and delineate some watersheds
using Whitebox tools. 




```{r, eval = F}

#File pointers
o_dem <- 'data/in/calhoun/CalhounBigDEM.tif'



# #Fill single cell pits (for hydrologic correctness)
# wbt_fill_single_cell_pits(o_dem,
#                           'data/out/calhoun/breach2.tif')

# #Breach depressions (better option that pit filling according to whitebox docu
# #mentation) The flat_increment bit was tricky for me and needed to be tuned.
# #### ----- WARNING Takes ~ 10 minutes with 18 cores ----- ####



check <- rast('data/in/calhoun/CalhounBigDEM.tif')  %>%
  crop(.,rast('data/in/calhoun/Bowl.tif'))


writeRaster(check,'data/in/test.tif', overwrite = T)

wbt_breach_depressions(o_dem,
                                  flat_increment = 0.1,
                       'data/out/calhoun/breached.tif')

#D8 pointer (what is a pointer? a flow direction grid? probably)
wbt_d8_pointer('data/out/calhoun/breached.tif',
               'data/out/calhoun/d8_pntr.tif')

#D8 flow
wbt_d8_flow_accumulation('data/out/calhoun/breached.tif',
                         'data/out/calhoun/d8_flow.tif',
                         out_type='catchment area')




# d8_max_1h <- rast('data/out/calhoun/d8_flow.tif')
# d8_max_1h[d8_max_1h > (1000*1000)] <- 0
# terra::writeRaster(d8_max_1k, 'data/out/calhoun/d8_max1k.tif')

wbt_slope(o_dem,
          'data/out/calhoun/slope.tif',
          units = 'percent')

wbt_slope('data/out/calhoun/breached.tif',
          'data/out/calhoun/breach_slope.tif',
          units = 'percent')

## MD infinity algorithm
wbt_md_inf_flow_accumulation('data/out/calhoun/breached.tif',
                             'data/out/calhoun/md_inf.tif',
                             out_type = 'catchment area')
#snap_points to outlets
wbt_snap_pour_points('data/in/calhoun/CalhounProjPoints.shp',
                     'data/out/calhoun/d8_flow.tif',
                     'data/out/calhoun/snapped_sites.shp',10)



#snap channel heads 20m
wbt_snap_pour_points('data/in/calhoun/CalhounChannelHeads.shp',
                     'data/out/calhoun/d8_flow.tif',
                     'data/out/calhoun/snapped_heads.shp',20)

# d8_pntr <- rast('data/out/calhoun/d8_pntr.tif')
# 
# 
# wbt_unnest_basins('data/out/calhoun/d8_pntr.tif',
#               'data/out/calhoun/big_snap.shp',
#               'data/out/calhoun/big_sheds.tif')
# 

#Watershed delineation as "whole watersheds'
wbt_unnest_basins('data/out/calhoun/d8_pntr.tif',
                  'data/out/calhoun/snapped_sites.shp',
                  'data/out/calhoun/cal_sheds.tif')


# Trim wasted area around delineated rasters
sheds_wide <- rast('data/out/calhoun/cal_sheds_1.tif') %>%
  trim(.)



terra::writeRaster(sheds_wide,'data/out/calhoun/cal_sheds.tif',
                   overwrite = T)
```


## Split watersheds into individual sheds

```{r, eval = F}
#Same order as read in (probably?)

sheds <- rast('data/out/calhoun/cal_sheds.tif')


pours <- st_read('data/in/calhoun/CalhounProjPoints.shp')

o_dem <- 'data/in/calhoun/CalhounDEM.tif'

#split sheds into subunits 
shed_splitter <- function(index){
  #Slow because domain is big
  single_shed = mask(sheds,sheds != index, maskvalue = 1) %>%
    trim(.)
  
  #Fast af, because crop and terra is awesome. 
  elev_shed <- crop(rast(o_dem),single_shed) %>%
    terra::mask(.,single_shed) %>%
    trim(.) 
  
  plot(elev_shed)

  #Write it out with assumed shed name. Can fix if not in order from snap
  terra::writeRaster(elev_shed,
                     filename = paste0('data/in/calhoun/',
                                      pours$id[index],
                                      '.tif'),
                     overwrite = T)
}

walk(1:4,shed_splitter)
```




## Whitebox super function

Build a function that does ~ 10 relevant analyses for a given DEM.  


```{r}
#Get the raw tifs that ends with ($) .tif

#Make a table for simpler names
#I'm ashamed of how manual this is. Was just easier.
dem_table <- tibble(files=c(list.files('data/in/coastal', 
                                     pattern = '.tif$', full.names=T),
                            'data/in/calhoun/Bowl.tif',
                            'data/in/calhoun/WS2.tif',
                            'data/in/calhoun/WS3.tif',
                            'data/in/calhoun/WS4.tif',
                            list.files('data/in/mtm',full.names=T)),
                    og_names = c(list.files('data/in/coastal',
                                          pattern = '.tif$'),
                                 'Bowl.tif',
                                 'WS2.tif',
                                 'WS3.tif',
                                 'WS4.tif',
                                 list.files('data/in/mtm')),
                    simple_names = c(
                                     'NY_ditched_a',
                                     'NY_ditched_b',
                                     'NY_naturalsih',
                                     'SC_natural_a',
                                     'SC_natural_b',
                                     'Bowl',
                                     'WS1',
                                     'Ws2',
                                     'WS3',
                                     list.files('data/in/mtm') %>%
                                       gsub('.tif','',.)),
                    treatment = c('novel','novel',
                                  'natural','natural','natural','natural',
                                  'novel','novel','novel',
                                  rep(c('novel','natural'),times=7))) 




super_whitebox <- function(files,og_names,simple_names,treatment){

  
  paster <- function(post = 'wbt_function'){
    paste0(out_folder,simple_names,post,'.tif')
  }
  
  #Name output folder
  out_folder <- gsub(og_names,'',files) %>%
    gsub('in','out',.)
  
  #Calculate ruggedness index
  wbt_ruggedness_index(dem = files,
                       output = paster('_ruggedness'))
  
  #Calculate releative elevation
  wbt_elev_relative_to_min_max(dem = files,
                               output = paster('_rel_elev'))
  
  #Edge density
  wbt_edge_density(files,
                   output = paster('_edge_density'),
                   norm_diff = 5)
  
  #surface area ratio 
  wbt_surface_area_ratio(files,
                         output = paster('_surface_area'))
  
  
  #Breach depressions (better option that pit filling according to whitebox docu
  #mentation) The flat_increment bit was tricky for me and needed to be tuned.
  wbt_breach_depressions(files,
                         output = paster('_breach'),
                         flat_increment=.1)
  
  
  #D8 flow
  wbt_d8_flow_accumulation(paster('_breach'),
                           paster('_d8_faa'),
                           out_type = 'catchment area')
  
  wbt_extract_streams(paster('_d8_faa'),
                      paster('_d8_strm5000'),
                      threshold = 5000)
                           
}


pwalk(dem_table,super_whitebox)
```





## Comparing other metrics (by hand... womp womp)


```{r}

reader <- function(r){
  v <- as.vector(rast(r)) %>% na.omit()
  place <- gsub(paste0('_',analysis,'.tif'),'',r)
  
  split_strings <- str_split_fixed(r,'/',n=4)
  
  
  table <- tibble(value = v,
                  name = analysis,
                  domain = split_strings[,3],
                  simple_names = gsub(paste0('_',analysis,'.tif'),'',split_strings[,4]))
}


comp_function <- function(analysis = 'ruggedness'){
  
  files <- list.files('data/out',
                      pattern = analysis,
                      recursive = T, 
                      full.names = T) 
  
    reader <- function(r){
      v <- as.vector(rast(r)) %>% na.omit() 
      v[v < -30] <- NA
      place <- gsub(paste0('_',analysis,'.tif'),'',r)
      
      split_strings <- str_split_fixed(r,'/',n=4)
      
      
      table <- tibble(value = v,
                      name = analysis,
                      domain = split_strings[,3],
                      simple_names = gsub(paste0('_',analysis,'.tif'),
                                          '',split_strings[,4]))
  }
  
  t <- map_dfr(files,reader) %>%
    inner_join(dem_table)
  
  # Way overkill to plot all data, but whatever. 
  gplot <- ggplot(t,aes(value, color = treatment, group = simple_names)) + 
    geom_density() + 
    facet_wrap(~domain,scales = 'free', ncol = 1 ) + 
    ggthemes::theme_few() + 
    ggtitle(analysis) + 
    scale_color_manual(name = '',values = c('green3','black')) + 
    theme(legend.position = c('top'), legend.direction = 'horizontal')
  
  
  summary <- t %>% 
    group_by(domain,treatment,simple_names) %>%
    summarize(median = median(value),
              mean = mean(value),
              min = min(value),
              max = max(value)) %>%
    mutate_if(is.numeric,round,2)
  
  
  return(list(gplot,summary))
}



```


### Ruggedness

```{r}
rug <- comp_function(analysis = 'ruggedness')

rug[[1]]

knitr::kable(rug[[2]])
```


### Edge density

Truly don't know what this means

```{r}

edge_dens <- comp_function('edge_density')

edge_dens[[1]]
```



### Elevation 

```{r}
rel_elev <- comp_function('breach')

rel_elev[[1]]

```


### FAA

No clue how to interpret in wetlands. Probably deeply flawed

```{r}
faa <- comp_function('d8_faa')

faa[[1]] + 
  scale_x_log10() 
```



### Surface area

Like edge density. Some kind of relative roughness? 

```{r}
sa <- comp_function('surface_area')

sa[[1]]
```


### Hypsometry 

This requires a different function.

#### Raw Elevation hypsometry
```{r}

files = dem_table$files[1]
og_names = dem_table$og_names[1]
simple_names = dem_table$simple_names[1]
treatment = dem_table$treatment[1]



hyp_reader <- function(files,og_names,simple_names,treatment){
    v <- as.vector(rast(files)) %>% na.omit() %>%
      round(.,1)
    
    unique_e <- unique(v)
    
    
    e.table <- tibble(elev = v) %>%
      mutate(e_bins = cut(elev,breaks=c(-Inf,unique_e,Inf))) %>% 
      group_by(e_bins) %>%
      summarize(count = n(),
                freq = 100*count/length(v),
                elev = first(elev) %>% na.omit(.)) %>%
      mutate(rel_elev = 100*((elev - min(elev))/(max(elev)-min(elev)))) %>%
      arrange(-elev) %>%
      mutate(area_above = cumsum(freq),
              simple_names = simple_names,
              treatment = treatment,
             domain = str_split_fixed(files,'/',4)[,3])

    
}

hyp_data <- pmap_dfr(dem_table,hyp_reader)


ggplot(hyp_data,aes(x=area_above, y = elev, color = treatment, group = simple_names)) + 
  geom_line() + 
  facet_wrap(~domain,nrow = 1, scales = 'free') + 
  ggthemes::theme_few() + 
  scale_color_manual(name = '',values = c('green3','black')) + 
  theme(legend.position = c('top'), legend.direction = 'horizontal')

```

#### Relative elevation hyp

```{r}
ggplot(hyp_data,aes(x=area_above, y = rel_elev, color = treatment, group = simple_names)) + 
  geom_line() + 
  facet_wrap(~domain,nrow = 1, scales = 'free') + 
  ggthemes::theme_few() + 
  scale_color_manual(name = '',values = c('green3','black')) + 
  theme(legend.position = c('top'), legend.direction = 'horizontal')

```


## Abandoned code that might be useful someday


```{r}
# #Transform all coastal to 2163 projection 
# transformer <- function(files,og_names,simple_names){
#   
#   #I don't know how to define this outside of the
#   #function without breaking the pmap functionality
#   # which is somewhat like a with() call here. 
#   paster <- function(post = 'wbt_function'){
#     paste0(out_folder,simple_names,post,'.tif')
#   }
#   
#   orig_proj <- rast(files)
#   
#     #Name output folder
#   out_folder <- gsub(og_names,'',files) %>%
#     gsub('in','out',.)
#   
# 
#   reproj <- project(orig_proj,'+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs') %>%
#     clamp(., lower = -15, upper = 15, values = NA)
#   
# 
#   terra::writeRaster(reproj,paster('_2163'),overwrite = T)
# }
# 
# 
# # I think I finally understand pmap on dfs
# # But this is actually a pwalk situation (no direct output, only side
# # effects (as in projecting and saving reprojected rasters))
# #Pwalk requires that all column names be options in the call
# #Makes sense? I guess? 
# pwalk(dem_table,transformer)
```

