---
title: "2_czo_ch"
author: "Matthew Ross"
date: "11/23/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}

library(sf)
library(raster)
library(terra)
library(leaflet)
library(stars)
library(leafem)
library(mapview)
library(tidyverse)
library(ggthemes)
library(rayshader)
library(patchwork)

#devtools::install_github('tylermorganwall/rayshader')
```


# Analyses of human-induced changes

## 3D Viz 


Rayshader super function

```{r}

```

### Calhoun

#### Bowl (natural)

```{r}



fac_colors <- colorRampPalette(colors=c('white','white','#00c37f','blue','purple4'))(20)




cal_viz <- function(dem_file = 'Bowl'){
  dem <- raster(paste0('data/in/calhoun/',dem_file,'.tif'))
  
  fac = raster(paste0('data/out/calhoun/md_inf.tif')) %>%
    crop(.,dem) %>%
    mask(., dem) %>%
    log10(.)
  
  
  
  dem_mat <- raster_to_matrix(dem)

  #Generate the flow accume overlay to drape over hillshade
  tempfilename = tempfile()
  png(tempfilename,width = nrow(dem_mat),height=ncol(dem_mat))
  par(mar = c(0,0,0,0),bg='transparent')
  raster::image((fac),
                axes = FALSE,xlab='',ylab='',
                col=fac_colors)
  
  dev.off()
  water_over = png::readPNG(tempfilename)
  

  
  rays <- ray_shade(dem_mat, zscale = 1,
                    sunaltitude = 45,
                    sunangle = 0)
  

  
  dem_mat %>%
    sphere_shade(texture = 'bw') %>% 
    add_overlay(water_over) %>%
    add_shadow(.,rays, max_darken = 0.5,
               rescale = F) %>%
    add_shadow(lamb_shade(dem_mat,
                          zscale = 1),
               0) %>%
    add_shadow(texture_shade(dem_mat, detail = 0.8,
                             contrast = 9,
                             brightness = 11),
               0) %>%
    plot_3d(dem_mat, zscale = 1)
  
}



cal_viz()
render_scalebar(limits = c(0,250),
                position = 'W',
                segments = 2,
                scale_length = c(0.25,0.75),
                label_unit = 'm',
                color_first = 'gray45',
                color_second = 'black')

```


#### WS2 (gullied)

```{r}

cal_viz('WS2')

```


### MTM

```{r}

mtm_viz <- function(dem_file = 'horse_post'){
  dem <- raster(paste0('data/in/mtm/',dem_file,'.tif'))
  
  fac <- raster(paste0('data/out/mtm/',dem_file,'_d8_faa.tif')) %>%
    log10(.)
  
  dem_mat <- raster_to_matrix(dem)
  
  
  #Generate the flow accume overlay to drape over hillshade
  tempfilename = tempfile()
  png(tempfilename,width = nrow(dem_mat),height=ncol(dem_mat))
  par(mar = c(0,0,0,0),bg='transparent')
  raster::image((fac),
                axes = FALSE,xlab='',ylab='',
                col=fac_colors)
  # raster::image(fliplr(shed_outlines),
  #               axes = FALSE,xlab='',ylab='',
  #               col=c('#EB008B','#0085ee'),add=T)
  # raster::image(fliplr(cross_outlines),
  #               axes=F,xlab='',ylab='',
  #               col=c('black'),add=T)
  dev.off()
  
  water_over = png::readPNG(tempfilename)
  
  dem_mat %>%
    sphere_shade(texture = 'bw') %>%
    add_overlay(water_over) %>%
    add_shadow(ray_shade(dem_mat, 
                         zscale = 5,
                         sunaltitude = 45,
                         sunangle = 0),
               max_darken = 0.5) %>%
    add_shadow(lamb_shade(dem_mat,
                          zscale = 5),
               0) %>%
    add_shadow(texture_shade(dem_mat, detail = 0.8,
                             contrast = 9,
                             brightness = 11),
               0) %>%
    plot_3d(dem_mat, zscale = 5)

}


mtm_viz('horse_post')
```


### Wetlands


```{r}
sc <- raster('data/in/coastal/SC_PostSandyDEM_2014_natural_a.tif')
ny <- raster('data/in/coastal/NY_PostSandyDEM_2014_ditched_a.tif')



wet_colors <- colorRampPalette(colors=c('white','white','#00c37f','blue','purple4'))(20)
par(bg='blue3')
plot(ny, col=rev(wet_colors), bg = 'blue')

dev.off()

```

### Site Map

```{r}

cals <- st_bbox(raster('data/in/calhoun/Bowl.tif')) %>% st_as_sfc(.) %>%
  st_transform(2163) %>%
  st_buffer(20000)
mtms <- st_bbox(raster('data/in/mtm/horse_pre.tif')) %>% st_as_sfc(.) %>%
  st_transform(2163)%>%
  st_buffer(20000)
sc_box <- st_bbox(sc) %>% st_as_sfc(.) %>%
  st_transform(2163)%>%
  st_buffer(20000)
ny_box <- st_bbox(ny) %>% st_as_sfc(.) %>%
  st_transform(2163)%>%
  st_buffer(20000)


sites <- c(cals,mtms,sc_box,ny_box) %>%
  st_buffer(20000) 

st_crs(sites) <- 2163
mapview(sites)


states <- USAboundaries::us_states() %>% 
  st_transform(2163) %>%
  .[st_bbox(sites) %>% st_as_sfc(.),] %>%
  filter(!stusps %in% c('OH','KY','TN','GA'))

sites
library(tmap)
tm_shape(states) + 
  tm_borders(col = 'black') +
tm_shape(cals) + 
  tm_borders(col = 'darkorange1') + 
tm_shape(mtms) + 
  tm_borders(col = 'blue') + 
tm_shape(ny_box) + 
  tm_borders(col = 'firebrick1') + 
tm_shape(sc_box) + 
  tm_borders(col = 'firebrick4') 
```


## Drainage density and proxies

3 Analyses for 3 landforms

- Channel outlet density for wetlands
- Channel head density for gullies
- Drainage density for MTM watersheds


### Calhoun data prep

```{r}


md_inf <- read_stars('data/out/calhoun/md_inf.tif')
slope <- read_stars('data/out/calhoun/slope.tif')

cal_snap <- st_read('data/out/calhoun/snapped_heads.shp') %>%
  st_transform(st_crs(md_inf)) %>%
  mutate(type = 'snapped')



cal_flow <- cal_snap %>%
  mutate(inf_uaa = st_extract(md_inf,cal_snap) %>% pull(1),
         slope = st_extract(slope,cal_snap) %>% pull(1))

```



### Calhoun Channel Head Slope vs Area




```{r}

cal_g <- ggplot(cal_flow, aes(x=inf_uaa/(1000*1000),y=slope,color=as.factor(Humans))) + 
  geom_point(size = 2) + 
  scale_x_log10(breaks = c(0.0001,0.001,0.01,0.1),
                labels = c('0.0001','0.001','0.01','0.1'),
                limits = c(0.0001,0.1)) + 
  scale_y_log10() + 
  theme_few() + 
  scale_color_manual(name = '',
                     values = c('black','red'),
                     labels = rev(c('novel',
                                'natural'))) + 
  xlab(expression(paste('Watershed Area (',m^2,')',))) + 
  ylab('Slope (degrees)') + 
  theme(legend.position = c(0.6,0.2)) + 
  labs(subtitle = 'Channel head locations') + 
  guides(color = 'none')



cal_g
```

## MTM Drainage Density (just 7 sites)


```{r}
facs <- list.files('data/out/mtm',
                   'faa.tif')

drain_length <- function(fac = facs[1]){
  fac_file <- paste0('data/out/mtm/',fac)
  
  uaa <- rast(fac_file) 
  area_km2 = max(values(uaa),na.rm = T)/(1000*1000)
  uaa[uaa < 8000] <- NA

  strm_length <- values(uaa) %>%
    .[!is.na(.)] %>%
    length(.)/1000
  tibble (area_km2,strm_length,
          site = str_split_fixed(fac,'_',3)[1],
          treat = str_split_fixed(fac,'_',3)[2])
}

t <- map_df(facs,drain_length)

mtm_g

mtm_g <- ggplot(t,aes(x=area_km2,
             y = strm_length,
             color = treat)) + 
  geom_point() + 
  scale_x_log10() + 
  scale_y_log10() + 
  ylab('Stream Length (km)') + 
  xlab(expression(paste('Watershed Area (k',m^2,')'))) + 
  theme_few() + 
  scale_color_manual(values = c('red','black'),
                     name = '',
                     labels = c('novel','natural')) +  
  theme(legend.position = c(0.2,0.8)) + 
  labs(subtitle = 'Drainage density') + 
  theme(legend.background = element_rect(fill = NA))

mtm_g
```



## Wetland outlet density

```{r}

outlets <- read_csv('data/in/coastal/wetland_density.csv') %>%
  filter(AreaKM < 4)








wet_out <- ggplot(outlets, aes(y = Number_Outlets,
                    x = AreaKM,
                    color = Altered)) + 
  geom_point() + 
  scale_color_manual(values = c('black','red'),
                     labels = c('natural','novel'),
                     name = '') + 
  theme_few() + 
  theme(legend.position = c(0.8,0.5)) + 
  ylab('# of outlets') + 
  xlab(expression(paste('Wetland area (k',m^2,')'))) + 
  labs(subtitle = 'Drainage outlet density') + 
  guides(color = 'none') 


```



## Combo

```{r}




Left <- cal_g / mtm_g / wet_out 

```
## Hyps

```{r}

dem_table <- tibble(files=c(list.files('data/in/coastal', 
                                     pattern = '.tif$', full.names=T),
                            'data/in/calhoun/Bowl.tif',
                            'data/in/calhoun/WS2.tif',
                            'data/in/calhoun/WS3.tif',
                            'data/in/calhoun/WS4.tif',
                            list.files('data/in/mtm',full.names=T)),
                    og_names = c(list.files('data/in/coastal',
                                          pattern = '.tif$'),
                                 'Bowl.tif',
                                 'WS2.tif',
                                 'WS3.tif',
                                 'WS4.tif',
                                 list.files('data/in/mtm')),
                    simple_names = c(
                                     'NY_ditched_a',
                                     'NY_ditched_b',
                                     'NY_naturalsih',
                                     'SC_natural_a',
                                     'SC_natural_b',
                                     'Bowl',
                                     'WS1',
                                     'Ws2',
                                     'WS3',
                                     list.files('data/in/mtm') %>%
                                       gsub('.tif','',.)),
                    treatment = c('novel','novel',
                                  'natural','natural','natural','natural',
                                  'novel','novel','novel',
                                  rep(c('novel','natural'),times=7))) 


files = dem_table$files[1]
og_names = dem_table$og_names[1]
simple_names = dem_table$simple_names[1]
treatment = dem_table$treatment[1]



hyp_reader <- function(files,og_names,simple_names,treatment){
    v <- as.vector(rast(files)) %>% na.omit() %>%
      round(.,1)
    
    unique_e <- unique(v)
    
    
    e.table <- tibble(elev = v) %>%
      mutate(e_bins = cut(elev,breaks=c(-Inf,unique_e,Inf))) %>% 
      group_by(e_bins) %>%
      summarize(count = n(),
                freq = 100*count/length(v),
                elev = first(elev) %>% na.omit(.)) %>%
      mutate(rel_elev = 100*((elev - min(elev))/(max(elev)-min(elev)))) %>%
      arrange(-elev) %>%
      mutate(area_above = cumsum(freq),
              simple_names = simple_names,
              treatment = treatment,
             domain = str_split_fixed(files,'/',4)[,3])

    
}

hyp_data <- pmap_dfr(dem_table,hyp_reader)

hypg<- ggplot(hyp_data,aes(x=area_above, y = rel_elev, color = treatment, group = simple_names)) + 
  geom_line() + 
  facet_wrap(~domain,nrow = 3, scales = 'free') + 
  ggthemes::theme_few() + 
  scale_color_manual(name = '',values = c('black','red')) + 
  theme(legend.position = c(0.4,0.4), legend.direction = 'horizontal') + 
  ylab('Relative elevation (%)') + 
  xlab('Area above (%)')



## Fix order. 
hypg
Left

hypg
```

